---
title: "Co-occurrence ACEs and diseases"
author: "Juan Rivillas"
date: "03/03/2023"
output:
  word_document: default
  html_document: default
---


#This R Marckdown included: i) Estimation adjusted Odd Ratios for confunders (ACES -> diseases)
ii) Construction sankey diagrams




```{r cars}
install.packages('table1')
library(table1)

install.packages('epiR') #Tools for the Analysis of Epidemiological Data
library(epiR)

install.packages('caret')
library(epiR)

install.packages('car')
library(car)

install.packages('compareGroups')
library(compareGroups)

#install.packages('InformationValue')
#library(InformationValue)

#help("package_version"="InformationValue")

install.packages ('networkD3')
library(networkD3)
```


##Step 1: Load the Data
#For this example, we’ll use the SABE aging population survey dataset from overall of biomarkers. 
```{r}
#load dataset
pathways_final

#view summary of dataset
summary(pathways_final)

#find total observations in dataset
nrow(pathways_final)
```



#Individual ACEs

Extract descriptive data
```{r}
table_2 <-table1 (~age + ethnic_group + emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + HTA + CVD + Diabetes + Obesity + ACE_score + Low_childhood_sep + Adult_sep | sex, data=pathways_final)

table_2
```



#co-occurence ACES
```{r}
test_compare2 <- compareGroups( data = pathways_final, 
               formula = emotional_abuse  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + Low_childhood_sep)

test_compare2 

test_table5 <- createTable( test_compare2, show.ci = TRUE )

test_table5
```
```


#ACEs score

#Create a new category for the ACEs variables.
```{r}
pathways_final$aces_score <- pathways_final$childhood_migration.x + pathways_final$emotional_abuse.x + pathways_final$poor_health2 + pathways_final$neglected_food + pathways_final$household_violence + pathways_final$early_infection.x



#create a vector aces score
pathways_final$aces_score <- factor(pathways_final$aces_score,
                levels = c("0","1","2","3","4","5","6"),
                labels = c("ACE 0","ACE 1", "ACE 2","ACE 3", "+ 4 ACE","+ 4 ACE", "+ 4 ACE"))

Desc(pathways_final$aces_score)
```

#Descriptive analysis of latentt groups

```{r}
#Characteristics ACEs
table_3 <-table1 (~age + sex + ethnic_group + area_residence.y + emotional_abuse + poor_food_environment + poor_childhood_health + early_infection + domestic_violence + childhood_migration  | ACE_score, data=pathways_final)
table_3

#Latent groups and outcomes
table_4 <-table1 (~age + sex + ethnic_group + HTA + CVD + Diabetes + Obesity + ACEs + Low_childhood_sep + Adult_sep | ACE_score, data=pathways_final)
table_4

```


#Correlation analysis

install.packages("compareGroups")
library("compareGroups")

```{r}
test_compare2 <- compareGroups( data = data_male, 
               formula = HTA  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare2 

test_table5 <- createTable( test_compare2, show.ci = TRUE )

test_table5
```

```{r}
test_compare3 <- compareGroups( data = pathways_final, 
               formula = CVD  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare3 

test_table3 <- createTable( test_compare3, show.ci = TRUE )

test_table3
```

```{r}
test_compare4 <- compareGroups( data = pathways_final, 
               formula = Diabetes  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare4 

test_table4 <- createTable( test_compare4, show.ci = TRUE )

test_table4
```


```{r}
test_compare1 <- compareGroups( data = pathways_final, 
               formula = Obesity  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare1 

test_table1 <- createTable( test_compare1, show.ci = TRUE )

test_table1
```

#Co-occurring ACE using corrlation matrix
```{r}
## as a default this function outputs a correlation matrix plot
ggcorrmat(
  data     = tetrachoric_correlation1,
  colors   = c("#B2182B", "white", "#4D4D4D"),
  title    = "Descriptive co-occurrence of childhood adversities ",
  subtitle = "Tetrachroric correlation using a correlalogram"
)
```



#This dataset contains the following information about 18044 individuals:
#We will use adversity in childhood and adult diseases to build a logistic regression model that predicts the probability that a given individual develop hypertension, obesity, cardiovascular diseases, or diabetes.

##Step 2: Create Training and Test Samples
Next, we’ll split the dataset into a training set to train the model on and a testing set to test the model on.

```{r}
#make this example reproducible
set.seed(1)

#Use 70% of SABE survey dataset as training set and remaining 30% as testing set
sample <- sample(c(TRUE, FALSE), nrow(pathways_final), replace=TRUE, prob=c(0.7,0.3))
train <- pathways_final[sample, ]
test <- pathways_final[!sample, ]  
```


##Step 3: Explore relationship between ACEs and diseases
```{r}
#Using TAB command
TAB <- table (pathways_final$ACEs, pathways_final$BMI)
TAB

barplot(TAB, beside = T, legend=T)

epi.2by2 (TAB, method="cross.sectional", conf.level=0.95)

1/1.04

```

#RLogistic regression by exposure-outcome-sex

```{r}
# Load the necessary libraries.
library(dplyr)
library(glm)
```

1. **Prepare your data.**
```{r}
# Split the data by sex.
data_male <- pathways_final %>% filter(sex == "Male")
data_female <- pathways_final %>% filter(sex == "Female")
```

2. Fit a logistic regression model for each sex.
```{r}
model_male <- glm(HTA ~ ACE_score +age, data = data_male, family = "binomial")
model_female <- glm(HTA ~ ACE_score, data = data_female, family = "binomial")
```

3. Predict the probability of disease for each sex.
```{r}
prediction_male <- predict(model_male, data_male, type = "response")
prediction_female <- predict(model_female, data_female, type = "response")
```

# Evaluate the performance of the models.
# ...

```{r}
prediction_male
```

# Interpret the results.
# ...






```{r}
m6 <- glm(
  HTA ~ ACE_score + age, data = data_male, family = binomial(logit)
)
summary(m6)
```


```{r}
# converting admit and rank
# columns into factor variables
pathways_final$HTA = as.factor(pathways_final$HTA)
pathways_final$Diabetes = as.factor(pathways_final$Diabetes)
pathways_final$CVD = as.factor(pathways_final$CVD)
pathways_final$Obesity = as.factor(pathways_final$Obesity)


odds_hta <- glm(HTA ~  ACEs + age + sex, data = pathways_final, family = binomial(logit))
odds_hta
```

Interpretation of the outcome (here displys more information): 
The coefficients in the output indicate the average change in log odds of hypertension. For example, not being exposed to adverse childhood is associated with an average decrease of -0.03396 in the log odds of hypertension.


```{r}
#fit logistic regression model (another way to build your models)
model <- glm(HTA ~  ACEs + age + sex + ethnic_group, family="binomial", data=pathways_final)

#disable scientific notation for model summary
options(scipen=999)

#view model summary
summary(model)
```

Interpretation of the outcome: 
The coefficients in the output indicate the average change in log odds of HTA. For example, i) being male is associated with an average decrease of -0.545886 in the log odds of HT: and ii) a one unit increase in age is associated with an average increase of 0.042667 in the log odds of hypertension.


The p-values in the output also give us an idea of how effective each predictor variable is at predicting the probability of developing hypertension:

P-value of aces:   0.736855
P-value of age:    <0.0000
P-value of gender: <0.0000  (male)
P-value of race:   0.00069 (indigenous)

We can see that age, gender and race seem to be important predictors since they have low p-values while adverse childhood is not nearly as important.

#Assessing Model Fit:
we can compute a metric known as McFadden’s R2, which ranges from 0 to just under 1. Values close to 0 indicate that the model has no predictive power. In practice, values over 0.40 indicate that a model fits the data very well.

We can compute McFadden’s R2 for our model using the pR2 function from the pscl package:

```{r}
pscl::pR2(model)["McFadden"]
```

A value of 0.02933238  is quite low for McFadden’s R2, which indicates that our model doesn't fits the data very well and hasn't high predictive power.

#Variable Importance:

We can also compute the importance of each predictor variable in the model by using the varImp function from the caret package:
We can also calculate the VIF values of each variable in the model to see if multicollinearity is a problem:
```{r}
#Contribution of each predictor
caret::varImp(model)

#calculate VIF values for each predictor variable in our model
car::vif(model)
```

Variable Importance (interpretation): Higher values indicate more importance. These results match up nicely with the p-values from the model. Age and gender male is by far the most important predictor variables, followed by ethnicity indigenous and then mestize.

VIF (interpretation): VIF values above 5 indicate severe multicollinearity. Since none of the  predictor variables in our models have a VIF over 5, we can assume that multicollinearity is not an issue in our model.


Step 4: Use the Model to Make Predictions
Once we’ve fit the logistic regression model, we can then use it to make predictions about whether or not an individual will develop hypertension based on their childhood adversity, age, gender and ethnicity:

```{r}
#define two individuals
new <- data.frame(aces = c("High-risk"), age =60, gender = c("female"), ethnic_group = c("white"))

#predict probability of defaulting
predict(model, new, type="response")
```

Interpretation of the outcome: 
The probability of an individual with 60 years old and afrocolombian female and exposed to high-risk of aces has a probability of developing hypertension of .6029637. Conversely, an white female with the age and exposed to aces has a probability of developing hta of 0.5084821. 

We can use the following code to calculate the probability of developing hypertension for every individual in our test dataset:

```{r}
#calculate probability of default for each individual in test dataset
predicted <- predict(model, test, type="response")
```

Step 5: Model Diagnostics (pending)

Another option summarising different models

```{r}
# full model
mod0 <- glm(hta ~ aces*age*gender*ethnic_group, data=dataBA2, family=binomial(link="logit"))
summary(mod0)

# full model minus four-way interaction
mod1 <- glm(hta ~ aces*age*gender*ethnic_group - aces:age:gender:ethnic_group, data=dataBA2, family=binomial(link="logit"))
summary(mod1)

# v1:v4 as only interaction
mod2 <- glm(hta ~ aces+age+gender+ethnic_group + aces:age:gender:ethnic_group, data=dataBA2, family=binomial(link="logit"))
summary(mod2)

anova(mod0, mod1, mod2)

# odds ratio coefficients and confidence intervals
library(MASS)
exp(cbind(coef(mod2), confint(mod2)))

```



#Addressing confounding using Mantel-Haenszel adjusted odds ratio
We’re concerned that the gender and race may confound the association between childhood adversity and hypertension in adulthood so we’ll stratify the data by race and gender and compute the Mantel-Haenszel adjusted odds ratio. As before, our tables can be prepared using either base R or tidyverse.

```{r}
#Stratified two by two table preparation using the table function in base R (ACES)
TAB_hta_gen  <- table (dataBA2$aces, dataBA2$hta, dataBA2$gender)
TAB_diabetes <- table (dataBA2$aces, dataBA2$diabetes,dataBA2$gender)
TAB_cvd      <- table (dataBA2$aces, dataBA2$cvd,dataBA2$gender)
TAB_obesity  <- table (dataBA2$aces, dataBA2$obesity, dataBA2$gender)

#Stratified two by two table preparation using the table function in base R (Childhood SEP)
TAB_hta_sep      <- table (dataBA2$low_childhood_sep, dataBA2$hta, dataBA2$gender)
TAB_diabetes_sep <- table (dataBA2$low_childhood_sep, dataBA2$diabetes, dataBA2$gender)
TAB_cvd_sep      <- table (dataBA2$low_childhood_sep, dataBA2$cvd, dataBA2$gender)
TAB_obesity_sep  <- table (dataBA2$low_childhood_sep, dataBA2$obesity, dataBA2$gender)

TAB_hta_race <- table (dataBA2$aces, dataBA2$hta, dataBA2$ethnic_group)

TAB_hta_gen
TAB_hta_race

```

ACES AND DISEASES

```{r}
test_compare4 <- compareGroups( data = pathways_final, 
               formula = Diabetes  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare4 

test_table4 <- createTable( test_compare4, show.ci = TRUE )

test_table4

```


```{r}
test_compare4 <- compareGroups( data = pathways_final, 
               formula = CVD  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare4 

test_table4 <- createTable( test_compare4, show.ci = TRUE )

test_table4

```


```{r}
test_compare4 <- compareGroups( data = pathways_final, 
               formula = HTA  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare4 

test_table4 <- createTable( test_compare4, show.ci = TRUE )

test_table4

```


```{r}
test_compare4 <- compareGroups( data = pathways_final, 
               formula = Obesity  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + ACEs + Low_childhood_sep + Adult_sep)

test_compare4 

test_table4 <- createTable( test_compare4, show.ci = TRUE )

test_table4

```


Compute the Mantel-Haenszel adjusted odds ratio for exposure and disease, adjusting for the effect of gender and race. Function epi.2by2 automatically calculates the Mantel-Haenszel odds ratio and risk ratio when presented with stratified contingency tables.

```{r}
#childhood adversity#
#Adjusted OR for gender hta
dat.epi05 <- epi.2by2(dat = TAB_hta_gen, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi05

#Adjusted OR for gender cvd
dat.epi06 <- epi.2by2(dat = TAB_cvd, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi06

#Adjusted OR for gender diabetes
dat.epi07 <- epi.2by2(dat = TAB_diabetes, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi07

#Adjusted OR for gender obesity
dat.epi08 <- epi.2by2(dat = TAB_obesity, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi08



#Low childhood SEP#
#Adjusted OR for gender hta
dat.epi09 <- epi.2by2(dat = TAB_hta_sep, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi09

#Adjusted OR for gender cvd
dat.epi10 <- epi.2by2(dat = TAB_cvd_sep, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi10

#Adjusted OR for gender diabetes
dat.epi11 <- epi.2by2(dat = TAB_diabetes_sep, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi11

#Adjusted OR for gender obesity
dat.epi12 <- epi.2by2(dat = TAB_obesity_sep, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi12

#tab_model(dat.epi12, file = "plot.html")


#Adjusted OR for race
dat.epi05 <- epi.2by2(dat = TAB_hta_race, method = "cross.sectional", 
   conf.level = 0.95, units = 100, interpret = FALSE, outcome = "as.columns")
dat.epi05
```
*Interpretation of the outcome (adjusted for gender): The Mantel-Haenszel test of homogeneity of the strata odds ratios is not significant (OR=0.051; p-value = 0.411). We accept the null hypothesis and conclude that the odds ratios for each strata of gender are the same. Because the stratum specific odds ratios are not significantly different the Mantel-Haenszel adjusted odds ratio provides an appropriate summary of the association between childhood adversity and hypertension.

After accounting for the confounding effect of gender, the odds of having a hypertension in later life for individuals exposed to adverse childhood experiences was 1.02 (95% CI 0.84 to 1.24) times that of non-exposed.



##################################################################################
#####################ANALYSIS REGRESSION ACES-DISEASES PAPER 2####################
##################################################################################
#option 3 logistic regression table and plot
Given their ease of calculation and simplicity, odds ratios are used near universally in medical reporting. While many researchers still prefer the tabular format to report ORs, the odds ratio plots, or forest plots, have recently come into vogue. There are many ways to generate OR plot in R, but the one that I find to be most intuitive and least onerous is the ‘finalfit’ package.

Plot the individual strata odds ratios and the Mantel-Haenszel summary odds ratio as an error bar plot to better understand how the Mantel-Haenszel adjusted odds ratio relates to the individual strata odds ratios.

Here is a quick demonstration:

First, let’s look at associations between our explanatory variable of interest (exposure) and other explanatory variables.

#1. Load the necessary libraries. for the analysis
```{r}
install.packages("finalfit")
install.packages("glm")
library(finalfit)
library(glm)

library(tidyverse)
library(finalfit)
library (dplyr)

# install.packages("devtools")
devtools::install_github("r-lib/conflicted")

#RLogistic regression by exposure-outcome-sex
library(dplyr)

library(glm)

#CochranArmitageTest {DescTools}
install.packages("DescTools")
library(DescTools)

```

#2. Prepare data and split it by sex (Explore sex differences)
```{r}
data_male <- pathways_final %>% filter(sex == "Male")
data_female <- pathways_final %>% filter(sex == "Female")
```

#3. Save two datasets males and females
```{r}
writexl::write_xlsx( x = data_male, path =  '/Users/macbookpro/Documents/PhD Project/Data/SABE/dataframes/data_male.xlsx' )

writexl::write_xlsx( x = data_female, path =  '/Users/macbookpro/Documents/PhD Project/Data/SABE/dataframes/data_female.xlsx' )
```


#4. Change to factor for the regressions 
```{r}
data_male$Obesity  = as.factor(data_male$Obesity)
data_male$HTA      = as.factor(data_male$HTA)
data_male$Diabetes = as.factor(data_male$Diabetes)
data_male$CVD      = as.factor(data_male$CVD)

data_male$sex2     = as.factor(data_male$sex2)
data_male$ACEs2    = as.factor(data_male$ACEs2)
data_male$ethnic_group2   = as.factor(data_male$ethnic_group2)
data_male$aces_score   = as.factor(data_male$aces_score)
data_male$ACE_score   = as.factor(data_male$ACE_score)

```


#5. Change the reference category using relevel((reference=advantage groups))  

```{r}
#What you can do is change it to an ordinary factor:
sex2 <- factor(dataBA2$gender, levels=c("female","male"))
ethnic_group2 <- factor(dataBA2$ethnic_group, levels=c("afrocolombian","mestize","white", "indigenous","other"))
ACEs2 <- factor(dataBA2$Childhood_adversity, levels=c("High-risk","Low-risk"))


Freq(pathways_final$ethnic_group2)

pathways_final$sex2 <- factor(pathways_final$sex,
                levels = c("Female","Male"),
                labels = c("2", "1"))

pathways_final$ACEs2 <- factor(pathways_final$ACEs,
                levels = c(Low-risk", "Moderate risk ACEs", "High-risk"),
                labels = c("1", "2", "3"))

pathways_final$ethnic_group2 <- factor(pathways_final$ethnic_group,
                levels = c("Afrocolombian","DK response", "Indigenous", "Mestize", "Other", "White"),
                labels = c("2","3","4","5","6","1"))


sex2 <- factor(dataBA2$sex2, levels=c("1","2"))
ethnic_group2 <- factor(dataBA2$ethnic_group2, levels=c("2","3", "4","5", "6","1"))
ACEs2 <- factor(dataBA2$Childhood_adversity, levels=c("1","2","3"))

#If we realized, we need to re-parametizer the model (putting always reference the wealthiest groups)
data_male <- within (data_male, ACEs2 <- relevel (ACEs2, ref = '1'))
data_male <- within (data_male, sex2 <- relevel (sex2, ref = '1'))
data_male <- within (data_male, ethnic_group2 <- relevel (ethnic_group2, ref = '1'))
data_male <- within (data_male, aces_score <- relevel (aces_score, ref = '1'))
data_male <- within (data_male, ACE_score <- relevel (ACE_score, ref = 'ACE 0'))

table(data_male$ACEs2)
table(data_male$sex2)
table(data_male$ethnic_group2)
table(data_male$aces_score)
```


#6. Specify explanatory variables of interest. Graphics OR for males

```{r}
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Diabetes"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```

#Specify explanatory variables of interest. Graphics OR for males
```{r}
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "CVD"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```

#Specify explanatory variables of interest. Graphics OR for males
```{r}
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "HTA"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```

#Specify explanatory variables of interest. Graphics OR for males
```{r}
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Obesity"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```

Perform a Cochran Armitage test for trend in binomial proportions across the levels of a single variable. This test is appropriate only when one variable has two levels and the other variable is ordinal. The two-level variable represents the response, and the other represents an explanatory variable with ordered levels. The null hypothesis is the hypothesis of no trend, which means that the binomial proportion is the same for all levels of the explanatory variable.

```{r}
tab8 <- table(data_male$ACE_score, data_male$Diabetes)
CochranArmitageTest(tab8)
```

```{r}
tab5 <- table(data_male$ACE_score, data_male$CVD)
CochranArmitageTest(tab5)
```

```{r}
tab6 <- table(data_male$ACE_score, data_male$HTA)
CochranArmitageTest(tab6)
```

```{r}
tab7 <- table(data_male$ACE_score, data_male$Obesity)
CochranArmitageTest(tab7)
```

Sensitivity analysis

m1 <- glm(HTA ~ ACE_score + age + ethnic_group,data = data_female, family = binomial(logit)
m2 <- glm(HTA ~ ACEs2 + age + ethnic_group,data = data_female, family = binomial(logit)

tab_model(m1)
```



#Graphics OR for females (we repeat same procedures)
```{r}
#Change to factor for the regressions
data_female$Obesity = as.factor(data_female$Obesity)
data_female$HTA      = as.factor(data_female$HTA)
data_female$Diabetes = as.factor(data_female$Diabetes)
data_female$CVD      = as.factor(data_female$CVD)

data_female$ACEs2    = as.factor(data_female$ACEs2)
data_female$ethnic_group2   = as.factor(data_female$ethnic_group2)
data_female$aces_score   = as.factor(data_female$aces_score)
data_female$ACE_score   = as.factor(data_female$ACE_score)
                         
                         
#If we realized, we need to re-parametizer the model (putting always reference the wealthiest groups)
data_female <- within (data_female, ACEs2 <- relevel (ACEs2, ref = '1'))
data_female <- within (data_female, ethnic_group2 <- relevel (ethnic_group2, ref = '1'))
data_female <- within (data_female, aces_score <- relevel (aces_score, ref = '1'))
data_female <- within (data_female, ACE_score <- relevel (ACE_score, ref = 'ACE 0'))
```

# Specify explanatory variables of interest
```{r}
iv <- c("ACE_score","age","ethnic_group2")
dv <- "Diabetes"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Specify explanatory variables of interest
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "CVD"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Specify explanatory variables of interest
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "HTA"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Specify explanatory variables of interest
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Obesity"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```



Perform a Cochran Armitage test for trend in binomial proportions across the levels of a single variable. The null hypothesis is the hypothesis of no trend, which means that the binomial proportion is the same for all levels of the explanatory variable.

```{r}
tab1 <- table(data_female$ACE_score, data_female$Diabetes)
CochranArmitageTest(tab1)
```

```{r}
tab2 <- table(data_female$ACE_score, data_female$CVD)
CochranArmitageTest(tab2)
```

```{r}
tab3 <- table(data_female$ACE_score, data_female$HTA)
CochranArmitageTest(tab3)
```

```{r}
tab4 <- table(data_female$ACE_score, data_female$Obesity)
CochranArmitageTest(tab4)
```


Not spliting dataframe by sex

```{r}
# Specify explanatory variables of interest all explanatory variables

iv <- c("ACEs2","age", "sex2", "ethnic_group2")
dv <- "Diabetes"

pathways_final %>% or_plot(dv, iv)

table3 <- pathways_final %>%
finalfit(dv, iv, 
           dependent_label_prefix = "")

table3
```

```{r}
# All explanatory variables of interest and four diseases

iv <- c("ACEs2","age","sex2", "ethnic_group2")
dv <- c("CVD")

pathways_final %>% or_plot(dv, iv)

table4 <- pathways_final %>%
finalfit(dv, iv, 
           dependent_label_prefix = "")

table4
```


```{r}
# All explanatory variables of interest and four diseases

iv <- c("ACEs2","age","sex2", "ethnic_group2")
dv <- c("Obesity")

pathways_final %>% or_plot(dv, iv)

table5 <- pathways_final  %>%
finalfit(dv, iv, 
           dependent_label_prefix = "")

table5
```




#Individual ACEs and diseases females
```{r}
#Change to factor for the regressions
data_female$emotional_abuse.x     = as.factor(data_female$emotional_abuse.x)
data_female$neglected_food     = as.factor(data_female$neglected_food)
data_female$household_violence     = as.factor(data_female$household_violence)
data_female$poor_health2     = as.factor(data_female$poor_health2)
data_female$early_infection.x       = as.factor(data_female$early_infection.x)
data_female$childhood_migration.x   = as.factor(data_female$childhood_migration.x)
data_female$Low_childhood_sep   = as.factor(data_female$Low_childhood_sep)
data_female$Adult_sep   = as.factor(data_female$Adult_sep)


#If we realized, we need to re-parametizer the model (putting always reference the wealthiest groups)
data_female <- within (data_female, emotional_abuse.x <- relevel (emotional_abuse.x, ref = '0'))
data_female <- within (data_female, neglected_food    <- relevel (neglected_food, ref = '0'))
data_female <- within (data_female, household_violence <- relevel (household_violence, ref = '0'))
data_female <- within (data_female, poor_health2 <- relevel (poor_health2, ref = '0'))
data_female <- within (data_female, early_infection.x <- relevel (early_infection.x, ref = '0'))
data_female <- within (data_female, childhood_migration.x <- relevel (childhood_migration.x, ref = '0'))

# Specify explanatory variables of interest
iv <- c("Adult_sep","age", "ethnic_group2")
dv <- "HTA"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```

#Individual ACEs and diseases males
```{r}
#Change to factor for the regressions
data_male$Obesity      = as.factor(data_male$Obesity)
data_male$HTA      = as.factor(data_male$HTA)
data_male$Diabetes = as.factor(data_male$Diabetes)
data_male$CVD      = as.factor(data_male$CVD)

data_male$ethnic_group2   = as.factor(data_male$ethnic_group2)

#Change to factor for the regressions
data_male$emotional_abuse.x     = as.factor(data_male$emotional_abuse.x)
data_male$neglected_food     = as.factor(data_male$neglected_food)
data_male$household_violence     = as.factor(data_male$household_violence)
data_male$poor_health2     = as.factor(data_male$poor_health2)
data_male$early_infection.x       = as.factor(data_male$early_infection.x)
data_male$childhood_migration.x   = as.factor(data_male$childhood_migration.x)
data_male$Low_childhood_sep   = as.factor(data_male$Low_childhood_sep)
data_male$Adult_sep   = as.factor(data_male$Adult_sep)


#If we realized, we need to re-parametizer the model (putting always reference the wealthiest groups)
data_male <- within (data_male, emotional_abuse.x <- relevel (emotional_abuse.x, ref = '0'))
data_male <- within (data_male, neglected_food    <- relevel (neglected_food, ref = '0'))
data_male <- within (data_male, household_violence <- relevel (household_violence, ref = '0'))
data_male <- within (data_male, poor_health2 <- relevel (poor_health2, ref = '0'))
data_male <- within (data_male, early_infection.x <- relevel (early_infection.x, ref = '0'))
data_male <- within (data_male, childhood_migration.x <- relevel (childhood_migration.x, ref = '0'))

# Specify explanatory variables of interest
iv <- c("Adult_sep","age", "ethnic_group2")
dv <- "HTA"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```



Validation Logistic regression with sjPlot package


```{r}
m3 <- glm(
  tot_sc_e ~ c160age + c12hour + c161sex + c172code, 
  data = efc,
  family = poisson(link = "log")
)

efc$neg_c_7d <- ifelse(efc$neg_c_7 < median(efc$neg_c_7, na.rm = TRUE), 0, 1)
m4 <- glm(
  neg_c_7d ~ c161sex + barthtot + c172code,
  data = efc,
  family = binomial(link = "logit")
)

tab_model(m3, m4)
```


```{r cars}

# load package
library(sjPlot)
library(sjmisc)
library(sjlabelled)


m6 <- glm(
  HTA ~ ACEs2 + age + ethnic_group2,
  data = data_male,
  family = binomial(logit)
)

m7 <- glm(
  CVD ~ ACEs2 + age + ethnic_group2,
  data = data_male,
  family = binomial(logit)
)

m8 <- glm(
  Diabetes ~ ACEs2 + age + ethnic_group2,
  data = data_male,
  family = binomial(logit)
)

m9 <- glm(
  Obesity ~ ACEs2 + age + ethnic_group2,
  data = data_male,
  family = binomial(logit)
)

tab_model(m6, m7, m8, m9)
```



Logistic models paper2
```{r}
tab_model(m, 
          show.reflvl = T, 
          show.intercept = F, 
          p.style = "numeric_stars")
```


second validation with ggstatplot package

```{r cars}
# for reproducibility
set.seed(123)

install.packages('ggstatsplot')

library(ggstatsplot)
library(lme4)
#> Loading required package: Matrix

# model object
mod <- lm(formula = HTA ~ ACEs + age + gender + ethnic_group, data = pathways_final)
mod1 <- lm(formula = Obesity ~ ACEs + age + sex + ethnic_group, data = pathways_final, contrasts = list(ACEs=c("Low-risk" =1))

# creating a plot
p <- ggcoefstats(mod1)

# looking at the plot
p

```



OD with ggstats

```{r cars}
install.packages("ggstatsplot")
library(ggstatsplot)
```



```{r cars}
model_obesity <- glm(
  obesity ~  aces,
  data = dataBA2,
  family = binomial(logit)
)

# looking at the plot
model_obesity
```


```{r cars}

# for reproducibility
set.seed(123)
library(lme4)

#> Loading required package: Matrix

# model object
mod <- lm(formula = obesity ~ aces * gender, data = dataBA2)

# creating a plot
p <- ggcoefstats(mod)

# looking at the plot
p
```


logistic regression with Bard - Machine learning

```{r cars}
1. Load the necessary libraries:
  
library(glmnet)

2. Load your data:
 
3. Split your data into training and test sets: 
set.seed(123)
train_index <- sample(1:nrow(data), size = 0.8 * nrow(data))
train_data <- data[train_index, ]
test_data <- data[-train_index, ]


Fit the logistic regression model:
4. model <- glm(HTA ~ ACEs - 1, family = "binomial", data = train_data)


5. Evaluate the model on the test set:
test_predictions <- predict(model, newdata = test_data, type = "response")
test_accuracy <- mean(test_predictions == test_data$HTA)
print(test_accuracy)


6. Interpret the results:

```


```{r cars}
glm(HTA ~ ACEs - 1, family = "binomial", )
```


###END ADJUSTED ODD RATIOS #####
