---
title: "Quantification of Biological Age BA algorithms"
author: "Juan Rivillas"
date: '2022-06-10'
output: html_document
---

#This code includes two steps
-Application published methods to generate BAA and compare them to one another using NHANEs datasets using BioAge package.
-Application published methods to generate BAA onto new test dataset (SABE).
-Quantification KDM-BA and HD onto SABE dataset and correlation BAA with one another.


Installation BioAge (via devtools):
```{r}
install.packages("tidyverse")
library(tidyverse)

install.packages("devtools")
devtools::install_github("dayoonkwon/BioAge")

install.packages("devtools")
install.packages("BioAge", dependencies=TRUE)
devtools::install_github("dayoonkwon/BioAge")
library(BioAge)

install.packages("BioAge", dependencies=TRUE)
remotes::install_github

install.packages("remotes")
remotes::install_github("dayoonkwon/BioAge", dependencies=TRUE)

install.packages("RNHANES")
install.packages("plyr")
library(RNHANES)

library(plyr)
library(dplyr) 
library(usethis)
library(devtools)
library(BioAge) 

```



@BIOAGE with set of markers

#The SABE dataset is loaded and has previously been downloaded and cleaned
```{r}
library(readxl)
dfbioage1 <- read_excel("Documents/PhD Project/Data/SABE/Dataframes/dfbioage1.xlsx")
View(dfbioage1) 
```


#Set of working markers (correlated) to compute BAA
```{r}
biomarkers = c('hdl','ldl','glucose','totchol','trig','hba1c','sbp', 'dbp', 'bmi','waist')
```


#Apply published methods to generate BAA and compare them to one another using NHANEs datasets. This is the first function dataset in the BioAge package.

Training HD algorithm using the NHANES 3 and projecting into NHANES 4 dataset. For this function, NHANES 3 included men and women who are between the ages of 20 and 30, and have observe biomarker data within clinically acceptable distributions (my reference of healthty population).

```{r}
#HD 
hd_nhanes = function(biomarkers) {

  #develop training dataset for HD
  train = NHANES3 %>%
    filter(age >= 20 & age <= 30 & pregnant == 0 & bmi < 30) %>%
    mutate(albumin = ifelse(albumin >= 3.5 & albumin <= 5, albumin, NA),
           albumin_gL = ifelse(is.na(albumin), NA, albumin_gL),
           alp = ifelse(gender == 2, ifelse(alp >= 37 & alp <= 98, alp, NA), ifelse(alp >= 45 & alp <= 115, alp, NA)),
           lnalp = ifelse(is.na(alp), NA, lnalp),
           bap = ifelse(gender == 2, ifelse(bap <= 14, bap, NA), ifelse(bap <= 20, bap, NA)),
           bun = ifelse(gender == 2, ifelse(bun >= 6 & bun <= 21, bun, NA), ifelse(bun >= 8 & bun <= 24, bun, NA)),
           lnbun = ifelse(is.na(bun), NA, lnbun),
           creat = ifelse(gender == 2, ifelse(creat >= 0.6 & creat <= 1.1, creat, NA), ifelse(creat >= 0.8 & creat <= 1.3, creat, NA)),
           creat_umol = ifelse(is.na(creat), NA, creat_umol),
           lncreat = ifelse(is.na(creat), NA, lncreat),
           lncreat_umol = ifelse(is.na(creat), NA, lncreat_umol),
           glucose = ifelse(glucose >= 60 & glucose <= 100, glucose, NA),
           glucose_mmol = ifelse(is.na(glucose), NA, glucose_mmol),
           glucose_fasting = ifelse(glucose_fasting >= 65 & glucose_fasting <= 110, glucose_fasting, NA),
           ttbl = ifelse(ttbl >= 0.1 & ttbl <= 1.4, ttbl, NA),
           uap = ifelse(gender == 2, ifelse(uap >= 2 & uap <= 7, uap, NA), ifelse(uap >= 2.1 & uap <= 8.5, uap, NA)),
           lnuap = ifelse(is.na(uap), NA, lnuap),
           basopa = ifelse(basopa >= 0 & basopa <= 2, basopa, NA),
           eosnpa = ifelse(eosnpa >=1 & eosnpa <= 7, eosnpa, NA),
           mcv = ifelse(gender == 2, ifelse(mcv >= 78 & mcv <= 101, mcv, NA), ifelse(mcv >= 82 & mcv <= 102, mcv, NA)),
           monopa = ifelse(monopa >= 3 & monopa <= 10, monopa, NA),
           neut = ifelse(neut >= 45 & neut <= 74, neut, NA),
           rbc = ifelse(gender == 2, ifelse(rbc >= 3.5 & rbc <= 5.5, rbc, NA), ifelse(rbc >= 4.2 & rbc <= 6.9, rbc, NA)),
           rdw = ifelse(rdw >= 11.5 & rdw <= 14.5, rdw, NA),
           cadmium = ifelse(cadmium >= 2.7 & cadmium <= 10.7, cadmium, NA),
           crp = ifelse(crp < 2, crp, NA),
           crp_cat = ifelse(is.na(crp), NA, crp_cat),
           lncrp = ifelse(is.na(crp), NA, lncrp),
           cyst = ifelse(cyst >= 0.51 & cyst <= 0.98, cyst, NA),
           ggt = ifelse(gender == 2, ifelse(ggt <= 37.79, ggt, NA), ifelse(ggt <= 55.19, ggt, NA)),
           insulin = ifelse(insulin >= 2.52 & insulin <= 24.1, insulin, NA),
           hba1c = ifelse(hba1c >= 4 & hba1c <= 5.6, hba1c, NA),
           lnhba1c = ifelse(is.na(hba1c), NA, lnhba1c),
           hdl = ifelse(gender == 2, ifelse(hdl >= 40 & hdl <= 86, hdl, NA), ifelse(hdl >= 35 & hdl <= 80, hdl, NA)),
           ldl = ifelse(ldl >= 80 & ldl <= 130, ldl, NA),
           trig = ifelse(trig >= 54 & trig <= 110, trig, NA),
           lymph = ifelse(lymph >= 20 & lymph <= 40, lymph, NA),
           wbc = ifelse(wbc >= 4.5 & wbc <= 11, wbc, NA),
           uap = ifelse(gender == 2, ifelse(uap >= 2.7 & uap <= 6.3, uap, NA), ifelse(uap >= 3.7 & uap <= 8, uap, NA)),
           sbp = ifelse(sbp < 120, sbp, NA),
           dbp = ifelse(dbp < 80, dbp, NA),
           meanbp = ifelse(meanbp < 93.33, meanbp, NA),
           pulse = ifelse(pulse >= 60 & pulse <= 100, pulse, NA),
           totchol = ifelse(totchol < 200, totchol, NA),
           fev = ifelse(fev >= mean(fev, na.rm = TRUE) * 0.8, fev, NA),
           fev_1000 = ifelse(is.na(fev), NA, fev_1000),
           vitaminA = ifelse(vitaminA >= 1.05 & vitaminA <= 2.27, vitaminA, NA),
           vitaminE = ifelse(vitaminE <= 28, vitaminE, NA),
           vitaminB12 = ifelse(vitaminB12 >= 100 & vitaminB12 <= 700, vitaminB12, NA),
           vitaminC = ifelse(vitaminC >= 23 & vitaminC <= 85, vitaminC, NA))

  #calculate hd
  hd_fem = hd_calc(data = NHANES4 %>%
                     filter(gender == 2),
                   reference = train %>%
                     filter(gender == 2),
                   biomarkers)

  hd_male = hd_calc(data = NHANES4 %>%
                     filter(gender == 1),
                   reference = train %>%
                     filter(gender == 1),
                   biomarkers)

  dat = rbind(hd_fem$data, hd_male$data)

  dat = left_join(NHANES4, dat[,c("sampleID","hd","hd_log")], by= "sampleID")
  fit = list(female = hd_fem$fit, male = hd_male$fit, nobs = hd_fem$fit$nobs + hd_male$fit$nobs)

  hd = list(data = dat, fit = fit)
  class(hd) = append(class(hd),'hd')
  return(hd)


}

```

#Training KDM Biological Age algorithm using the NHANES 3 and projecting into NHANES 4 dataset.

```{r}
#KDM
kdm_nhanes = function (biomarkers) {

  #calculate training KDM
  train_fem = kdm_calc(data = NHANES3 %>%
                         filter(age >= 30 & age <= 75 & pregnant == 0, gender == 2),
                       biomarkers, fit = NULL, s_ba2 = NULL)
  train_male = kdm_calc(data = NHANES3 %>%
                          filter(age >= 30 & age <= 75 & pregnant == 0, gender == 1),
                        biomarkers, fit = NULL, s_ba2 = NULL)

#calculate test modified KDM
  test_fem = kdm_calc(data = NHANES4 %>%
                        filter(gender == 2),
                      biomarkers, fit = train_fem$fit, s_ba2 = train_fem$fit$s_ba2)
  test_male = kdm_calc(data = NHANES4 %>%
                         filter(gender == 1),
                       biomarkers, fit = train_male$fit, s_ba2 = train_male$fit$s_ba2)

#combine calculated kdm
  test = rbind(test_fem$data, test_male$data)

#combine data
  dat = left_join(NHANES4, test[,c("sampleID", "kdm", "kdm_advance")], by = "sampleID")
  fit = list(female = train_fem$fit, male = train_male$fit, nobs = test_fem$fit$nobs + test_male$fit$nobs)

  kdm = list(data = dat, fit = fit)
  class(kdm) = append(class(kdm), "kdm")
  return(kdm)

}
```


#Apply published methods to generate BAA onto new test dataset (SABE). This is the second function dataset in the BioAge package.

For HD algorithm., the constructed variable is based on a mahalanobis distance statistic, which is theoretically the distance between observations and a hypothetically healthy, young cohort. Here, we train separately for men and women who are between the ages of 20 and 30 and not pregnant, and have observe biomarkers data within clinically acceptable distributions.

```{r}
hd_calc = function(data=dfbioage1, reference=train, biomarkers= c('hdl', 'ldl','glucose','totchol','trig','hba1c', 'sbp', 'dbp', 'bmi', 'waist')){

  ref = as.matrix(reference[, biomarkers])
  dat = as.matrix(data[, biomarkers])

  #standardize variables by mean and sd of reference population
  for (j in 1:ncol(dat)){
    dat[,j] <- (dat[,j] - mean(ref[,j], na.rm = TRUE)) / sd(ref[,j], na.rm = TRUE)
  }

  for (j in 1:ncol(ref)){
    ref[,j] <- (ref[,j] - mean(ref[,j], na.rm = TRUE)) / sd(ref[,j], na.rm = TRUE)
  }

  dat = na.omit(dat)
  ref = na.omit(ref)

  if(nrow(ref) == 1){
    warning("The reference matrix must have more than one row")
  }

  else {

    means = colMeans(ref)
    cv_mat = var(ref)

  }

  if(nrow(dat) == 1){
    warning("The function does not work with single-row data")
  }

  else {

    dat = as.matrix(dat)
    hd = rep(NA, nrow(dat))

    for (x in 1:nrow(dat)){
      hd[x] <- sqrt((dat[x,] - means) %*% solve(cv_mat) %*% (dat[x,] - means))
    }

  }

  dat = data %>%
    select(sampleID,all_of(biomarkers)) %>%
    na.omit()

  dat$hd = hd/sd(hd)
  dat$hd_log = log(hd)/sd(log(hd))
  nobs = sum(!is.na(dat$hd))

  dat = left_join(data, dat[,c("sampleID","hd","hd_log")], by = "sampleID")
  fit = list(mcov = means, cov_mat = cv_mat, nobs = nobs)
  hd = list(data = as.data.frame(dat),fit = fit)

  class(hd) = append(class(hd),'hd')
  return(hd)

}
```

#Projecting KDM algorithm onto new data.

```{r}
kdm_calc = function (data=dfbioage1, biomarkers= c('hdl','ldl','glucose','totchol','trig','hba1c', 'sbp', 'dbp', 'bmi', 'waist'), fit = NULL, s_ba2 = NULL) {

  dat = data
  bm = biomarkers
  bm_dat = t(select(dat, bm))

  rm(biomarkers)

  if (is.null(fit)) {

    lm_age = t(apply(bm_dat, 1, function(x) {
      sm = summary(lm(x ~ dat$age))
      return(c(sm$sigma, sm$r.sq, sm$coef[1,1], sm$coef[2,]))

    }))

    colnames(lm_age) <- c("RMSE", "r_squared", "B_intercept", "B_age", "se(B)_age", "T_val_age", "pval_age")
    lm_age <- as.data.frame(lm_age)

    lm_age$r1 = abs((lm_age$B_age / lm_age$RMSE) * sqrt(lm_age$r_squared))
    lm_age$r2 = abs((lm_age$B_age / lm_age$RMSE))
    lm_age$n2 = (lm_age$B_age / lm_age$RMSE) * (lm_age$B_age / lm_age$RMSE)

    age_range = range(dat$age, na.rm = TRUE)
    rchar = sum(lm_age$r1) / sum(lm_age$r2)
    s_r = ((1 - (rchar * rchar)) / (rchar * rchar) * (((age_range[2] - age_range[1]) ^ 2) / (nrow(lm_age) * 12)))

  }

  else {

    lm_age = fit$lm_age
    s_r = fit$s_r

  }
  #end dat conditional

  n1 = bm_dat
  for (r in 1:nrow(n1)) {

    x = rownames(n1)[r]
    n1[r,] = ((bm_dat[x,] - lm_age[x,"B_intercept"]) * (lm_age[x,"B_age"] / (lm_age[x,"RMSE"] * lm_age[x,"RMSE"])))

  }

  rm(r); rm(x)

  BAe_n = apply(n1, 2, sum); rm(n1)
  BAe_d = sum(lm_age$n2)

  BAe = BAe_n / BAe_d
  BA_CA = BAe - dat$age

  s2 = sd(BA_CA, na.rm = TRUE) ^ 2
  nobs = sum(!is.na(BA_CA))

  if (is.null(s_ba2)) {

    s_ba2 = s2 - s_r

  }

  else {

    s_ba2 = s_ba2

  }

  dat$kdm = ((BAe_n) + (dat$age / (s_ba2))) / ((BAe_d) + (1 / (s_ba2)))
  dat$kdm_advance = dat$kdm - dat$age

  fit = list(lm_age = lm_age, s_r = s_r, s_ba2 = s_ba2, s2 = s2, nobs = nobs)

  kdm = list(data = as.data.frame(dat), fit = fit)
  class(kdm) = append(class(kdm), "kdm")
  return(kdm)

}
```


#train algorithms in NHANES 3 and project biological aging measures in NHANES 3

```{r}
#HD using NHANES (separate training for men and women)
hd = hd_calc(data= dfbioage1, reference = NHANES3, biomarkers=c('hdl','ldl','glucose','totchol','trig','hba1c', 'sbp', 'dbp', 'bmi', 'waist'))

#Extract HD dataset
data =hd$data


#KDM bioage using NHANES (separate training for men and women)
kdm = kdm_calc(data= dfbioage1, biomarkers=c('hdl','ldl','glucose','totchol','trig','hba1c', 'sbp', 'sbp', 'bmi', 'waist'), fit = NULL, s_ba2 = NULL)


#extract KDM dataset
data=kdm$data

```


#compare original KDM-BA and HD algorithms with algorithms composed with new set of biomarkers
```{r}
#assemble SABE dataset with projected biological aging measures for analysis
data = merge(hd$data, kdm$data)
View(data)
```

#Now, save new data with four BAA.
```{r}
#Save data
writexl::write_xlsx( x =data, path =  '/Users/macbookpro/Documents/PhD Project/Data/SABE/dataframes/dataBA.xlsx' )
```


#Plot Pearson correlation BBA and CA
```{r}
#select biological age variables
agevar = c("kdm","kdm_advance","hd","hd_log")

#prepare labels
label = c("KDM\nBiological Age",
          "Modified-KDM\nBiological Age",
          "Homeostatic\nDysregulation",
          "Log\nHomeostatic\nDysregulation")

#plot age vs bioage
plot_ba(data, agevar, label)
```

#correlation matriz BAA
```{r}
#select biological age variables
agevar = c("kdm","kdm_advance","hd","hd_log")

#prepare lables
#values should be formatted for displaying along diagonal of the plot
#names should be used to match variables and order is preserved
label = c(
  "kdm"="KDM\nBiological Age\nAdvancement",
  "kdm_advance"="Modified-KDM\nBiological Age\nAdvancement",
  "hd" = "Homeostatic\nDysregulation",
  "hd_log" = "Log\nHomeostatic\nDysregulation")

#use variable name to define the axis type ("int" or "float")
axis_type = c(
  "kdm"="float",
  "kdm_advance"="float",
  "hd"="flot",
  "hd_log"="float")

#plot BAA corplot
plot_baa(data, agevar, label, axis_type)
```




```{r}
corr_cross(corr, type = 2)
corr_cross(corr_aces, type = 2)
corr_cross(corr_sep, type = 2)
corr_cross(corr_ba, type = 2)
```



#compare classes and bio age measures
```{r}
table_bioage_cg <- compareGroups( data, 
               formula = Class ~ age + kdm + kdm_advance + hd + hd_log)

table_bioage_ci <- createTable(x = table_bioage_cg, show.ci = T, show.n = T, show.all = T )

table_bioage_ci

export2word( x = table_bioage_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bioage_ci.docx')
```


compare classes and bio age measures
```{r}
table_bachild_cg <- compareGroups( data, 
               formula = migration_yo ~ age + kdm + kdm_advance + hd + hd_log)

table_bachild_ci <- createTable(x = table_bachild_cg, show.ci = T, show.n = T, show.all = T )

table_bachild_ci

export2word( x = table_bachild_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bachild_ci.docx')
```


```{r}
#education
table_baedu_cg <- compareGroups( data, 
               formula = education ~ age + kdm + kdm_advance + hd + hd_log)

table_baedu_ci <- createTable(x = table_baedu_cg, show.ci = T, show.n = T, show.all = T )

table_baedu_ci

export2word( x = table_baedu_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_baedu_ci.docx')
```


```{r}
#occupation
table_baocu_cg <- compareGroups( data, 
               formula = sep_occupation ~ age + kdm + kdm_advance + hd + hd_log)

table_baocu_ci <- createTable(x = table_baocu_cg, show.ci = T, show.n = T, show.all = T )

table_baocu_ci

export2word( x = table_baocu_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_baocu_ci.docx')
```


```{r}
#current sep
table_basep_cg <- compareGroups( data, 
               formula = current_sep ~ age + kdm + kdm_advance + hd + hd_log)

table_basep_ci <- createTable(x = table_basep_cg, show.ci = T, show.n = T, show.all = T )

table_basep_ci

export2word( x = table_basep_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_basep_ci.docx')
```


```{r}
#health insurance
table_bahi_cg <- compareGroups( data, 
               formula = health_insurance ~ age + kdm + kdm_advance + hd + hd_log)

table_bahi_ci <- createTable(x = table_bahi_cg, show.ci = T, show.n = T, show.all = T )

table_bahi_ci

export2word( x = table_bahi_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bahi_ci.docx')
```


```{r}
#social policy
table_bacash_cg <- compareGroups( data, 
               formula = social.cash.transfer ~ age + kdm + kdm_advance + hd + hd_log)

table_bacash_ci <- createTable(x = table_bacash_cg, show.ci = T, show.n = T, show.all = T )

table_bacash_ci

export2word( x = table_bacash_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bacash_ci.docx')
```


```{r}
#social policy
table_banut_cg <- compareGroups( data, 
               formula = nutrition_program ~ age + kdm + kdm_advance + hd + hd_log)

table_banut_ci <- createTable(x = table_banut_cg, show.ci = T, show.n = T, show.all = T )

table_banut_ci

export2word( x = table_banut_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_banut_ci.docx')
```



```{r}
#sex
table_basex_cg <- compareGroups( data, 
               formula = sex ~ age + kdm + kdm_advance + hd + hd_log)

table_basex_ci <- createTable(x = table_basex_cg, show.ci = T, show.n = T, show.all = T )

table_basex_ci

export2word( x = table_basex_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_basex_ci.docx')
```



```{r}
#residence
table_bares_cg <- compareGroups( data, 
               formula = residence ~ age + kdm + kdm_advance + hd + hd_log)

table_bares_ci <- createTable(x = table_bares_cg, show.ci = T, show.n = T, show.all = T )

table_bares_ci

export2word( x = table_bares_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bares_ci.docx')
```



```{r}
#early infection
table_bainf_cg <- compareGroups( data, 
               formula = raised_bmi ~ age + kdm + kdm_advance + hd + hd_log)

table_bainf_ci <- createTable(x = table_bainf_cg, show.ci = T, show.n = T, show.all = T )

table_bainf_ci

export2word( x = table_bainf_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bainf_ci.docx')
```



```{r}
#early infection
table_bainf_cg <- compareGroups( data, 
               formula = raised_bmi ~ age + kdm + kdm_advance + hd + hd_log)

table_bainf_ci <- createTable(x = table_bainf_cg, show.ci = T, show.n = T, show.all = T )

table_bainf_ci

export2word( x = table_bainf_ci, file = '/Users/macbookpro/Documents/PhD Project/Data/SABE/Outputs/table_bainf_ci.docx')
```





