---
title: "Code 3 visualisation data and hypothesis testing"
author: "Juan Rivillas"
date: "04/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code 2. Bivariate analysis and statistical testing.
Note: Please note that exploratory data analysis (mainly tables) are reported in code 2. This code reports additional findings using visualization data.

Three main steps are described in this markdown.

**Step 1. Import dataframe (biomarkers.imp2)**
        1.1. Working directory and loading of packages required for analysis and visualisation data.
        1.2. Importing of new transformed data base.
        1.3. Verifying upload and exploring original data base.
        
**Step 2. Basic statistics: Basic calculations, plotting charts to review spreading and distribution of data**.
        2.1. Describe distribution the data (plots)
        2.2. Summarise the data (Mean, Median, Mode): Outliers, errors, missing data
        2.3. Measure the variability in the data (inter-quartile range, variance, SD, FWHM): Variation and covariation
        2.4. Quantify the strength of associations between variables (Pearson correlation coefficient)
             Asking questions about data (correlation matrices)
             Clustering

## Step 1. Preparing work space (use new transformated data base).

**1.1. Working directory and loading of packages required for analysis and visualisation data**.

Assign file pathway to save automatically plots and outcomes in pdf format.

```{r}
setwd("/Users/macbookpro/Documents/PhD Project/Data/SABE/Markdowns")
```

Install CRAN packages for visusalisation data.

```{r}
install.packages(c("tidyverse", "colorspace", "corrr",  "cowplot",
                   "ggdark", "ggforce", "ggrepel", "ggridges", "ggsci",
                    "ggtext", "ggthemes", "grid", "gridExtra", "patchwork",
                    "rcartocolor", "scico", "showtext", "shiny",
                    "plotly", "highcharter", "echarts4r"))
 
## install from GitHub since not on CRAN
#devtools::install_github("JohnCoene/charter")
```
Retrieve palettes using various methods listed below. 

```{r}
install.packages("MetBrewer")
library(MetBrewer)
install.packages("devtools")
devtools::install_github("BlakeRMills/MetBrewer")
met.brewer("Hiroshige")

["Hiroshige",["#aadce0","#ef8a47","#376795","#f7aa58","#72bcd5","#ffe6b7","#e76254","#1e466e","#ffd06f","#528fad"]]
```

**1.2. Importing of new data base (Loading data EXCEL files)**.

```{r}
library(readxl)
dfsabe <- read_excel("~/Documents/PhD Project/Data/SABE/Dataframes/biomarkers.imp2.xlsx")
View(dfsabe)
```

Another option for CSV files:
dfsabe <- read_csv("biomarkers.imp2.csv")

**1.3. Verifying upload and exploring imputed data base**.

```{r }
skimr::skim(dfsabe)
```


# Step 2. Basic statistics: Basic calculations, plotting charts to review spreading and distribution of data.

2.1. Distributions.
Use a Distributions widget to look at histograms for each column.
Which variables appear to be normally distributed? What happens when the data are split by type?


```{r}
#Metabolic system
plot(dfsabe$triglycerides)
plot(dfsabe$glucose)
plot(dfsabe$ldl)
plot(dfsabe$hdl)
plot(dfsabe$total_cholesterol)
plot(dfsabe$glycated_haemoglobin)

#cardiometabolic system
plot(dfsabe$dyastolic_bprh_2)
plot(dfsabe$systolic_bprh_2)
plot(dfsabe$bmi_2)
plot(dfsabe$wc)

plot(dfsabe$age)

```



ggplot(dfsabe, aes(ldl, age_group, color=sex))+geom_point()
ggplot(biomarkers, aes(total_cholesterol, age, color=sex))+geom_point()
ggplot(biomarkers, aes(bmi, age, color=sex))+geom_point()
ggplot(biomarkers, aes(hdl, age, color=acesum))+geom_point()
ggplot(biomarkers, aes(glucose, age, color=educational_level))+geom_point()

**Cardiometabolic biomarkers and adverse childhood experiences**

```{r}
g <- ggplot(dfsabe, aes(x = triglycerides, y = acesum, color = age)) +
  geom_point() +
  labs(x = "Triglycerides", y = "Adverse childhood experiences", color = "age")

g +
  geom_segment(
    aes(x = triglycerides, xend = triglycerides,
        y = acesum, yend = acesum),
    size = 0.8
  ) +
  geom_hline(aes(yintercept = acesum), color = "gray70", size = 0.6) +
  geom_jitter(size = 2, alpha = 0.25, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 5)

```

```{r}
g <- ggplot(dfsabe, aes(x = glycated_haemoglobin, y = acesum, color = sex)) +
  geom_point() +
  labs(x = "Glycated haemoglobin", y = "Adverse childhood experiences", color = "sex")

g +
  geom_segment(
    aes(x = glycated_haemoglobin, xend = glycated_haemoglobin,
        y = acesum, yend = acesum),
    size = 0.8
  ) +
  geom_hline(aes(yintercept = acesum), color = "gray70", size = 0.6) +
  geom_jitter(size = 2, alpha = 0.25, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 5)
```


```{r}
g <- ggplot(dfsabe, aes(x = bmi_2, y = acesum, color = sex)) +
  geom_point() +
  labs(x = "BMI", y = "Adverse childhood experiences", color = "sex")

g +
  geom_segment(
    aes(x = bmi_2, xend = bmi_2,
        y = acesum, yend = acesum),
    size = 0.8
  ) +
  geom_hline(aes(yintercept = acesum), color = "gray70", size = 0.6) +
  geom_jitter(size = 2, alpha = 0.25, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 5)
```



```{r}
g <- ggplot(dfsabe, aes(x = dyastolic_bprh_2, y = acesum, color = age)) +
  geom_point() +
  labs(x = "DBP", y = "Adverse childhood experiences", color = "age")

g +
  geom_segment(
    aes(x = dyastolic_bprh_2, xend = dyastolic_bprh_2,
        y = acesum, yend = acesum),
    size = 0.8
  ) +
  geom_hline(aes(yintercept = acesum), color = "gray70", size = 0.6) +
  geom_jitter(size = 2, alpha = 0.25, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 5)
```


```{r}
ggplot(dfsabe, aes(x = sex, y = triglycerides)) +
  geom_boxplot(fill = "grey92") +
  ## use either geom_point() or geom_jitter()
  geom_point(
    ## draw bigger points
    size = 2,
    ## add some transparency
    alpha = .3,
    ## add some jittering
    position = position_jitter(
      ## control randomness and range of jitter
      seed = 1, width = .2
    )
  )
```


**Cardiometabolic biomarkers and family socioeconomic position**

```{r}
g <- ggplot(dfsabe, aes(x = triglycerides, y = family_sep, color = age)) +
  geom_point() +
  labs(x = "Triglycerides", y = "Family socioeconomic position", color = "age")

g +
  geom_segment(
    aes(x = triglycerides, xend = triglycerides,
        y = family_sep, yend = family_sep),
    size = 0.8
  ) +
  geom_hline(aes(yintercept = family_sep), color = "gray70", size = 0.6) +
  geom_jitter(size = 2, alpha = 0.25, width = 0.2) +
  stat_summary(fun = mean, geom = "point", size = 5)

```







hist(dfsabe$triglycerides, colors = met_brew(name="Hiroshige", n=123, brew_type="continuous"))

met.brewer(name, n, type = c("discrete", "continuous"))


library(ggplot2)
ggplot(data=dfsabe, aes(x=age, y=glucose, fill=sex)) +
geom_violin() +
scale_fill_manual(values=met.brewer("Hiroshige", 3))




import met_palettes


c("Mean", "Median", "Minimum", "Maximum"), lwd = 5)




Biomarkers by sex








library("ggdist")

ggplot(biomarkers, aes(x = sex, y = ldl)) + 
  ggdist::stat_halfeye(
    adjust = .5, 
    width = .6, 
    .width = 0, 
    justification = -.3, 
    point_colour = NA) + 
  geom_boxplot(
    width = .25, 
    outlier.shape = NA
  ) +
  geom_point(
    size = 1.3,
    alpha = .3,
    position = position_jitter(
      seed = 1, width = .1
    )
  ) + 
  coord_cartesian(xlim = c(1.2, NA), clip = "off")

ggplot(biomarkers, aes(x = ldl, y = HbA1c)) + 
  ggdist::stat_halfeye(
    adjust = .5, 
    width = .6, 
    .width = 0, 
    justification = -.3, 
    point_colour = NA) + 
  geom_boxplot(
    width = .25, 
    outlier.shape = NA
  ) +
  geom_point(
    size = 1.3,
    alpha = .3,
    position = position_jitter(
      seed = 1, width = .1
    )
  ) + 
  coord_cartesian(xlim = c(1.2, NA), clip = "off")
  













```{r} 
# examine the distribution of bmi
hist(biomarkers$high.waist.circumference, breaks = 10)
```

hist(biomarkers$glucose, breaks = 5) 

```{r} 
# examine the distribution of bmi
hist(biomarkers$bmi)
```

First, examine the distribution of your data, to make appropriate cut-points.

```{r} 
# examine the distribution of bmi
hist(biomarkers$bmi, breaks = 10)
```

```{r} 
summary(biomarkers$bmi, na.rm=T)
```





#color checking: The package has been updated to check for colorblind-friendlyness You can list out the colorblind-friendly palettes with the following code
MetBrewer::colorblind_palettes

met.brewer(name="Hiroshige",n=10,type="discrete")

ggplot(data=biomarkers, aes(x=age, y=glucose, color=age)) +
  geom_point(size=1) +
  scale_color_gradientn (colors=met.brewer("Hiroshige"))
  
ggplot(data = biomarkers,
       mapping = aes(x = sex,
                     y = ldl,
                     color = sex)) +
   geom_jitter()+
   geom_boxplot() 

ggplot(data = biomarkers,
       mapping = aes(x = sex,
                     y = hdl,
                     color = sex)) +
   geom_jitter()+
   geom_boxplot() 
   
ggplot(data = biomarkers,
       mapping = aes(x = sex,
                     y = glucose,
                     color = sex)) +
   geom_jitter()+
   geom_boxplot()  

ggplot(data = biomarkers,
      mapping = aes(x = triglycerides,
                    y = age,
                   color = as.factor(sex))) +
  geom_point(alpha = 0.5) 
  
  ggplot(data = biomarkers,
      mapping = aes(x = triglycerides,
                    y = total_cholesterol,
                   color = as.factor(sex))) +
  geom_point(alpha = 0.5)
 
 ggplot(data = biomarkers,
      mapping = aes(x = ldl,
                    y = hdl,
                   color = sex)) +
  geom_point(alpha = 0.5)
 
#Assign plot to a variable ABEJAS!

cholesterol_plot <- ggplot(data = biomarkers,
                       aes(x = ldl,
                           y = hdl))
                           
 cholesterol_plot +
 geom_hex()
 
 ---
 hdl_plot <- ggplot(data = biomarkers,
                       aes(x = hdl,
                           y = total_cholesterol))
                           
 hdl_plot +
 geom_hex() 
 
 ---
 cholesterol2_plot <- ggplot(data = biomarkers,
                       aes(x = triglycerides,
                           y = total_cholesterol))
                           
 cholesterol2_plot +
 geom_hex()
 
 
#examples plots
ggplot(data=biomarkers, aes(x=sex, y=ldl, fill=sex)) +
  geom_violin() +
  scale_fill_manual(values=met.brewer("Hiroshige", 3))
  
  ggplot(data=biomarkers, aes(x=waist, y=weight, color=waist)) +
  geom_point(size=2) +
  scale_color_manual(values=met.brewer("Hiroshige", 3))

#display discrete data key characteristics by bar graphs and charts

sex<-ggplot(biomarkers, aes(x=sex, y=sex, fill=sex)) +
  geom_bar(stat="identity")+theme_minimal()
sex + scale_color_continuous("#e76254", "#ef8a47")

sex<-ggplot(biomarkers, aes(x=sex, y=sex, fill=sex)) +
  geom_bar(stat="identity")+theme_minimal()
sex + scale_color_continuous()


"Hiroshige", 3


education<-ggplot(biomarkers, aes(x=educational_level, y=sex, fill=educational_level)) +
  geom_bar(stat="identity" + theme_minimal()
education + scale_color_continuous()

barplot(biomarkers$sex)


#The :: is called namespace and can be used to access a function without loading the package. Here, you could also run library(readr) first and chic <- read_csv(...) afterwards.
tibble::glimpse(biomarkers)

gb <- ggplot(biomarkers, aes(x = triglycerides, y = total_cholesterol, color = age)) +
  geom_point() +
  labs(x = "triglycerides", y = "total_cholesterol", color = "age")

gb + scale_color_continuous()

gb <- ggplot(biomarkers, aes(x = triglycerides, y = total_cholesterol, color = sex)) +
  geom_point() +
  labs(x = "triglycerides", y = "total_cholesterol", color = "sex")

gb + scale_color_continuous()








Box plots.
Use the Box Plot widget to compare variable distributions between types.
Which variables have significantly different means between types?



#Plotting

```{r}
biomarkers %>%                            # begin with dataframe
  count(sex, family_sep) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = family_sep,                # map outcome to x-axis
        fill = sex,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(sex, sep_occupation) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = sep_occupation,                # map outcome to x-axis
        fill = sex,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(sex, current_sep) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = current_sep,                # map outcome to x-axis
        fill = sex,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(sex, early_infection) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = early_infection,                # map outcome to x-axis
        fill = sex,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(sex, acesum) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = acesum,                # map outcome to x-axis
        fill = sex,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(ethnic_group, acesum) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = acesum,                # map outcome to x-axis
        fill = ethnic_group,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```




```{r}
biomarkers %>%                            # begin with dataframe
  count(age_group, acesum) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = acesum,                # map outcome to x-axis
        fill = age_group,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```



```{r}
biomarkers %>%                            # begin with dataframe
  count(age_group, medicationsum) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = medicationsum,                # map outcome to x-axis
        fill = age_group,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(age_group, comorbidities) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = comorbidities,                # map outcome to x-axis
        fill = age_group,                 # map age to the fill
        y = n))                           # map the counts column `n` to the height
```

```{r}
biomarkers %>%                            # begin with dataframe
  count(sex, comorbidities) %>%     # group and tabulate counts by two columns
  ggplot()+                               # pass new data frame to ggplot
    geom_col(                             # create bar plot
      mapping = aes(   
        x = comorbidities,                # map outcome to x-axis
        fill = sex,                 # map sex to the fill
        y = n))                           # map the counts column `n` to the height
```



```{r}
compareGroups(group ~ age + sex + smoke + waist + hormo, data = predimed, 
    selec = list(hormo = sex == "Female", waist = waist > 20))
```



Histograms of a biomarkers variables
```{r}
hist(dfsabe$triglycerides)
hist(dfsabe$glucose)
hist(dfsabe$ldl)
hist(dfsabe$hdl)
hist(dfsabe$total_cholesterol)
hist(dfsabe$glycated_haemoglobin)
hist(dfsabe$dyastolic_bprh_2)
hist(dfsabe$systolic_bprh_2)
hist(dfsabe$bmi_2)
hist(dfsabe$wc)

hist(dfsabe$age)
```




Box plots

#Violin plots, similar to box plots except you’re using a kernel density to show where you have the most data, are a useful visualization

#Start with box plots

g<-ggplot(biomarkers, aes(x=HbA1c, y=family_sep))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=family_sep),
      position = position_jitter(width = 0.1))+coord_flip()

#example with SABE
g<-ggplot(SABE_Colombia, aes(x=HbA1c, y=family_sep))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=family_sep),
      position = position_jitter(width = 0.1))+coord_flip()
      
g<-ggplot(SABE_Colombia, aes(x=HbA1c, y=ldl))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=family_sep),
      position = position_jitter(width = 0.1))+coord_flip()
      
      _____
g<-ggplot(biomarkers, aes(x=HbA1c, y=educational_level))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=educational_level),
      position = position_jitter(width = 0.1))+coord_flip()


g<-ggplot(biomarkers, aes(x=HbA1c, y=acesum))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=HbA1c),
      position = position_jitter(width = 0.1))+coord_flip()
      
      
g<-ggplot(biomarkers, aes(x=hdl, y=acesum))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=hdl),
      position = position_jitter(width = 0.1))+coord_flip()      


g<-ggplot(biomarkers, aes(x=ldl, y=acesum))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=ldl),
      position = position_jitter(width = 0.1))+coord_flip()

      
g<-ggplot(biomarkers, aes(x=total_cholesterol, y=sep_occupation))
g+geom_boxplot(fill="darkseagreen4")

g+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=total_cholesterol),
      position = position_jitter(width = 0.1))+coord_flip()

      
hdl<-ggplot(biomarkers, aes(x=total_cholesterol, y=current_sep))
hdl+geom_boxplot(fill="darkseagreen4")

hdl+geom_violin(alpha=0.5, color="gray")+geom_jitter(alpha=0.5, aes(color=current_sep),
      position = position_jitter(width = 0.1))+coord_flip()



#Improved historgram adding data points
ggplot(data, aes(x = group, y = value)) +
  geom_boxplot(fill = "grey92") +
  ## use either geom_point() or geom_jitter()
  geom_point(
    ## draw bigger points
    size = 2,
    ## add some transparency
    alpha = .3,
    ## add some jittering
    position = position_jitter(
      ## control randomness and range of jitter
      seed = 1, width = .2
    )
  )



Scatter plots.
Add a Scatter Plot widget to look at covariation between variables.
Use Find Informative Projections to find the pairs of variables that maximise separation between types.
Use a Correlations widget to suggest other interesting projections of the data.



**************
NOTE: DONT RUN!!!

Loading library for multiple scatter plot with marginal density plots
```{r}
install.packages("ggpubr")
library(ggpubr)
```



```{r}
# Scatter plot colored by groups ("ACEs")
aces1 <- ggscatter(dfsabe, x = "total_cholesterol", y = "acesum",
                color = "age_group", palette = "jco",
                size = 3, alpha = 0.6)+
  border()                                         

# Marginal density plot of x (top panel) and y (right panel)
xplot <- ggdensity(dfsabe, "total_cholesterol", fill = "age_group",
                   palette = "jco")
yplot <- ggdensity(dfsabe, "acesum", fill = "age_group", 
                   palette = "jco")+
  rotate()

# Cleaning the plots
yplot <- yplot + clean_theme() 
xplot <- xplot + clean_theme()

# Arranging the plot
#ggarrange(xplot, NULL, sp, yplot, 
          #ncol = 2, nrow = 2,  align = "hv", 
          #widths = c(2, 1), heights = c(1, 2),
          #common.legend = TRUE)

```
**************


Scatter plot with marginal density plots




K-mean clustering
Clustering is an unsupervised task that tries to divide the dataset into subgroups that contain similar data points. It can
be a useful element of exploratory data analysis.
k-Means clustering operates directly on the feature space. This means that we do not need to compute distances between
data points.
k-means is a heuristic method, which means that it may not always produce the same result.

Task
Use a k-Means widget to find clusters in the data set.
How many clusters appear to be present?
Which features can be used to separate clusters?
How do the clusters correspond to type?
---

# Inter quartile range 
quantile(biomarkers$ldl)

ggboxplot(SABE_Colombia, y = " HEMOGLOBINA ", width = 0.5)

xtabs(~ESTRATO_2 + P141 + P121, SABE_Colombia = SEP_HIS_sex)

xtabs(~ESTRATO_2 + P141 + P121, data = SABE_Colombia)
