---
title: 'Early-life socioeconomic inequality, adverse childhood experiences and biological aging in Colombia  1_imputation data'
author: "Juan Rivillas"
date: "28/02/2022"
output:
  html_document:
    df_print: paged
  word_document: 
    fig_width: 4
    fig_height: 3
    fig_caption: yes
  pdf_document: default
---

## Contents Code 1: Data processing and preparation for analysis.
Code 1 is divided into three main steps:

Step 1. Preparing work space **(data manipulation steps)**.
Step 2. **Imputation missing values** for data cleaning and exploration.
Step 3. **Tidying data**: organize data in a consistent data structure.

Step 1 and 2 is described in this markdown. 

Step 1. Preparing work space **(data manipulation steps)**. This step includes:
        1.1. Working directory and loading of packages required for the tidying data.
        1.2. Importing of original data base.
        1.3. Verifying upload and exploring original data base.
        1.4. Subset individuals with biomarkers -> filter BIOMARKER ==1 (Selection of 4,092 observations).
        1.5. Verifying upload and exploring individual socioeconomic and biomarkers data (selection of 54 columns).
        1.6. Overview of the entire dataframe (missing values, complete rates, histograms).

Step 2. **Imputation missing values** for data cleaning and exploration. This step includes:
        2.1. Loading packages to work with missing values.
        2.2. Descriptive statistics with missing values.
        2.3. Understanding and exploring the missing value patterns.
        2.4. Imputating missing values using multiple imputations with mice package.
        2.5. Inspecting the quality of the imputed values.
        
---

## Step 1. Preparing work space (data manipulation steps).

**1.1. Working directory and loading of packages required for the tidying data**.

Assign file pathway to save automatically plots and outcomes in pdf format.

```{r}
setwd("/Users/macbookpro/Documents/PhD Project/Data/SABE/Markdowns")
```

Install CRAN packages and install the relevant libraries. This code chunk shows the loading of packages required for the analyses.

```{r eval=FALSE, include=FALSE}
install.packages(c("dplyr", "ggplot2", "vcdExtra","epiDisplay", "data.table", "formattable", "table1", "haven", "devtools"))

#Loading the relevant libraries into the work space to tidying and transformation data.

library(tidyverse)  #Tidy data and transformation
library(dplyr)       
library(epiDisplay) #to make frequency table in r (nicer version)
library(DescTools)  #single and cumulative frequencies values are reported.
library(table1)     #to Create HTML Tables of Descriptive Statistics in epi
library("ggplot2")  #improved plots and bar charts
library(haven)
library(devtools)
```

**1.2. Importing of original SABE data base (Loading data EXCEL files)**.
Note: All columns containind individual information were deleted to avoid any potential identification of individuals. 

```{r}
library(readxl)
SABE_Colombia <- read_excel("~/Documents/PhD Project/Data/SABE/Dataframes/SABE_Colombia.xlsx")
```
Another option for CSV files:
SABE_Colombia <- read_csv("SABE_Colombia.csv")

**1.3. Verifying upload and exploring original data base**.
Here, Glimpse makes it possible to see every column within data frame. It's a little like str applied to a data frame but it tries to show as much data as possible.

```{r}
str(SABE_Colombia)
dim(SABE_Colombia)
glimpse(SABE_Colombia)
head(SABE_Colombia, 50)
```

**1.4. subset individuals with biomarkers** -> filter BIOMARKER ==1 (Selection of 4,092 observations).
There are 22,639 individuals and 1,139 columns in the original database. The column "Biomarkers" was filtered to select individuals with collected biomarkers data (4,092 observations).

```{r}
newdata <- SABE_Colombia[ which(SABE_Colombia$biomarkers=='1'), ]
```

Selection of 52 sociodemographic, socioeconomic, and health outcomes variables present in the data.

```{r}
myvars <- c("ID", "current_sep", "sep_occupation", "family_sep","area_residence","educational_level","health_insurance","neglected_food", "household_violence", "migration", "migration_yo","emotional_abuse","poor_health2",
"early_infection1","early_infection2","early_infection3","early_infection4","early_infection5","early_infection6","early_infection7","hdl", "ldl", "total_cholesterol", "triglycerides", "glucose", "glycated_haemoglobin", "weight", "height", "wc","systolic_bprh","dyastolic_bprh","systolic_bplh", "dyastolic_bplh", "hta", "diabetes", "cvd", "medication","P835","P835_1","P835_2","P835_3","P835_4","P835_5","P835_6","P835_7","age", "sex", "ethnic_group", "alcohol", "smoking","cash_transfer1", "cash_transfer2")
```

**1.5. Verifying upload and exploring individual socioeconomic and biomarker data**.

```{r}
biomarkers <- newdata[myvars]
```

**1.6. Overview of the entire dataframe** (missing values, complete rates, histograms).
A first glance there are 9 character variables and 43 numeric variables. There are 15 variables with not missing value, there are 34 variables with 20% missing values, and there are five columns with 50% missing values (medication and diastolic and systolic blood pressure in the dataframe).

```{r}
skimr::skim(biomarkers)
```

## Step 2. Imputation missing values for data cleaning and exploration.

**2.1. Loading packages to work with missing values**.

```{r}
library(mice)   #Loading the mice package
library(VIM)    #Loading the following package for looking at the missing values
library(lattice)
library (naniar)#tidy ways to summarise, visualize, and manipulate misisng data.
```

**2.2. Descriptive statistics with missing values**.
To visualise the missings for each variable.

```{r Descriptive statistics with missing values (n)}
gg_miss_var(biomarkers)
```

str(biomarkers) #to see every column within data frame
Look at the percent missing by setting and set the y limits to be between 0 and 100.

```{r Descriptive statistics with missing values (%), eval=FALSE, include=FALSE}
gg_miss_var(biomarkers, 
            show_pct = TRUE) + 
  ylim(0, 100)
```
look at the missings across cases.

```{r Description of number of missings across cases}
gg_miss_case(biomarkers) 
```

Plot variables sorted by number of missing and individuals with missings values to identify and visualize patterns of missing entries.

```{r Description of proportion of missingness and missingness pattern}
biomarkers_miss = aggr(biomarkers, col=mdc(1:2), numbers=TRUE, sortVars=TRUE, labels=names(biomarkers), cex.axis=.7, gap=3, ylab=c("Proportion of missingness","Missingness Pattern"))
```

**2.3. Understanding and exploring the missingness patterns**.
This graph check association between the value or a variable and the missingness of another one using plots. 

```{r}
matrixplot(biomarkers, sortby = 2)
```

Here, we apply additional helpful visual approach (box plot) to draw margin plot of missing data patterns. For instance, the red box plot on the left shows the distribution of age with systolic blood pressure missing at the bottom of the graph.
If the red and blue box plots are very similar, then the assumption of the MCAR data is correct. This means the probability of being missing is the same for all cases, then the data are said to be missing completely at random (MCAR). This approach was applied to six combination of variables with missing data. Overall, missing data is distributed similarly observations in the data.

```{r}
marginplot(biomarkers[, c("age", "ethnic_group")], col = mdc(1:2), cex.numbers = 1.2, pch = 19)
marginplot(biomarkers[, c("wc", "family_sep")], col = mdc(1:2), cex.numbers = 1.2, pch = 19)
marginplot(biomarkers[, c("sep_occupation", "dyastolic_bprh")], col = mdc(1:2), cex.numbers = 1.2, pch = 19)
marginplot(biomarkers[, c("age", "weight")], col = mdc(1:2), cex.numbers = 1.2, pch = 19)
marginplot(biomarkers[, c("glucose", "early_infection1")], col = mdc(1:2), cex.numbers = 1.2, pch = 19)
marginplot(biomarkers[, c("age","systolic_bprh")], col = mdc(1:2), cex.numbers = 1.2, pch = 19)
```

```{r}
#summary with missing values
pct_miss(biomarkers)          #percentage of missing value in the data
n_miss(biomarkers)            #number of missing values 
n_complete(biomarkers)        #without missing value
n_miss(biomarkers$migration_yo) #number of missing value for var
```

as_shadow(biomarkers)         #a matrix with missing and non missing
bind_shadow(biomarkers)       #Initial matrix concatenated with the matrix with missing and non missing

Check percentage of missing data patterns in each column.
There are 79% observations with not missing values. There are 21% observations with missing values in medication variables, dyastolic and systolic blood pressure, early-life infection and childhood experiences variables (which corresponds 42,841 missing values).

Apply clustering to find similar missingness groups by setting. 

```{r}
vis_miss(biomarkers, sort_miss = TRUE)
```

There are a lot of different clusters here, to see clear relationships here we apply another technique to try arranging by different variables using arrange. It is possibly to identify common missigness patterns: there are same 16% missing values in family sep, early-life infection, childhood experiences variables, and ethnic groups (n=686).

```{r Cluster missingness group 1: family sep, eval=FALSE, include=FALSE}
biomarkers %>%
  arrange(family_sep) %>%
  vis_miss()
```

Four measurement variables of dyastolic and systolic blood pressure have the same number of individuals with missing values (51%).

```{r Cluster of missingness group 2: blood pressure, eval=FALSE, include=FALSE}
biomarkers %>%
  arrange(systolic_bprh) %>%
  vis_miss()
```

```{r Cluster of missingness group 3: medication variables, eval=FALSE, include=FALSE}
biomarkers %>%
  arrange(P835_2) %>%
  vis_miss()
```

Check the combination and patterns of missingness by looking at an upset plot of the missingness.
Here, medication has the most missing values and there are 2684 cases where five medications have missing values together.

```{r Description combination and patterns of missingness by intersection size}
gg_miss_upset(biomarkers)
```

Look at histogram to clearly depicts the influence of missing values in the variables checking intersection sizes across variables.

```{r Combination and patterns of missingness all variables by intersection size}
n_var_miss(biomarkers)
gg_miss_upset(biomarkers, nsets = n_var_miss(biomarkers))
```

Looking at the missingness patterns by age missing values.

```{r Description missingness patterns by age missings}
gg_miss_fct(biomarkers, age)
```

```{r Description missingness patterns by ldl}
gg_miss_fct(biomarkers, ldl)
```

```{r Description missingness patterns by total cholesterol}
gg_miss_fct(biomarkers, total_cholesterol)
```

Explore the missingness by some categorical variable using facetting to take a closer look at the distribution of missing values.

```{r Distribution of missing values of age and systolic blood pressure by family SEP categories, eval=FALSE, include=FALSE}
ggplot(biomarkers, 
       aes(x = age, 
           y = systolic_bprh)) + 
  geom_miss_point() + 
  facet_wrap(~biomarkers$family_sep) + scale_fill_brewer(palette =)
```

```{r Distribution of missing values of age and systolic blood pressure by neglected food categories, eval=FALSE, include=FALSE}
ggplot(biomarkers, 
       aes(x = age, 
           y = systolic_bprh)) + 
  geom_miss_point() + 
  facet_wrap(~biomarkers$neglected_food) + scale_fill_brewer(palette =)
```

```{r Distribution of missing values of age of migratio and weight by occupation status categories, eval=FALSE, include=FALSE}
ggplot(biomarkers, 
       aes(x = migration_yo, 
           y = weight)) + 
  geom_miss_point() + 
  facet_wrap(~biomarkers$sep_occupation)
```

```{r Distribution of missing values of age of migration and dyastolic blood pressure by poor health categories, eval=FALSE, include=FALSE}
ggplot(biomarkers, 
       aes(x = migration_yo, 
           y = dyastolic_bplh)) + 
  geom_miss_point() + 
  facet_wrap(~biomarkers$poor_health2) 
```

Reviewing the cumulative sum of missing values, reading the rows of the dataframe from the top to bottom using a plot.

```{r Cummulative sum of missing values}
gg_miss_case_cumsum(biomarkers)
```

```{r eval=FALSE, include=FALSE}
biomarkers %>%
  bind_shadow() %>%
  ggplot(aes(x = age,
             fill = systolic_bprh)) +
  geom_density()
```

**2.4. imputing missing values using** using predictive mean matching as imputation method with mice package (Generates Multivariate Imputations by Chained Equations MICE). 

Note:This step takes 60 minutes processing data.

Here is an explanation of the parameters used:

m  – Refers to 5 imputed data sets.
maxit – Refers to no. of iterations taken to impute missing values.
method – Refers to method used in imputation. we used predictive mean matching.

```{r eval=FALSE, include=FALSE}
tempData <- mice(biomarkers,m=5,maxit=10, meth='pmm',seed=) 
```

Check the imputation method used for each variable (mice makes it very easy to do).
```{r eval=FALSE, include=FALSE}
tempData$meth
```

Check the imputed data for each observation (first column left) within each imputed dataframe (first row at the top).
The missing values have been replaced with the imputed values in the first of the five dataframe.

tempData$imp$family_sep
tempData$imp$ethnic_group
tempData$imp$systolic_bprh
tempData$imp$systolic_bplh
tempData$imp$dyastolic_bprh
tempData$imp$dyastolic_bplh

Since there are 5 imputed data sets, we select any using complete() function to export imputed data.
After imputation, there are 90% observations with missing any information and 10% observation with missing values (which corresponds 21,999 missing values).

```{r eval=FALSE, include=FALSE}
biomarkers.imp <- complete(tempData,1)

#summary with missing values
pct_miss(biomarkers.imp)          #percentage of missing value in the data
n_miss(biomarkers.imp)            #number of missing values 
n_complete(biomarkers.imp)        #without missing value
```

**2.5. Inspecting the quality of imputed data**.
Inspecting the distribution of original and imputed data and the quality of the imputations using the xyplot() and densityplot() functions to verify imputations. First, using a scatterplot and plot age against all the other variables. Second, check whether that the shape of the magenta points (imputed) matches the shape of the blue ones (observed) using density plots. The matching shape tells us that the imputed values are indeed “plausible values”.

```{r eval=FALSE, include=FALSE}
xyplot(tempData,age ~ family_sep + sep_occupation, pch=18, cex=1)
xyplot(tempData,age ~ early_infection1 + early_infection2 + early_infection3 + early_infection4 + early_infection5, early_infection6, pch=18, cex=1)
xyplot(tempData,age ~ emotional_abuse + neglected_food + household_violence + poor_health2, migration, pch=18, cex=1)
xyplot(tempData,age ~ ethnic_group, pch=18, cex=1)
xyplot(tempData, age  ~ weight + height + wc, pch=18, cex=1)
xyplot(tempData,migration_yo ~ weight + height + wc, pch=18, cex=1)
xyplot(tempData, dyastolic_bprh ~ emotional_abuse + neglected_food + household_violence + poor_health2, pch=18,cex=1)
xyplot(tempData, migration_yo ~ dyastolic_bprh + systolic_bprh, pch=18,cex=1)
```

densityplot(tempData)
densityplot(tempData$family_sep)
densityplot(tempData$early_infection2)
densityplot(tempData$neglected_food)
densityplot(tempData$dyastolic_bprh)
densityplot(tempData$systolic_bprh)
densityplot(tempData$weight)
densityplot(tempData$height)
densityplot(tempData$ethnic_group)

Save imputed data base for transformation data.
```{r eval=FALSE, include=FALSE}
writexl::write_xlsx( x = biomarkers.imp, path =  '/Users/macbookpro/Documents/PhD Project/Data/SABE/dataframes/biomarkers.imp.xlsx' )
```
