---
title: "Regression tables BA SEP and predictor models"
author: "Juan Rivillas"
date: "06/07/2022"
output: html_document
---

Load package
```{r}
library(ggplot2)
library(ggstatsplot)
```

Check effects in the prediction based on regressions models applying DAG 
```{r}
set.seed(123)

## models with exposures-mediators
mod_ba_child  <- stats::lm(formula = kdm_advance ~ age*sex + Class + childhood_sep + household_violence + migration_yo, data = dataBA)
mod_ba_aces   <- stats::lm(formula = kdm_advance ~ age*sex + household_violence + migration_yo, data = dataBA)
mod_ba_adult  <- stats::lm(formula = kdm_advance ~ age*sex + education + sep_occupation + current_sep, data = dataBA)

ggcoefstats(mod_ba_child)
ggcoefstats(mod_ba_aces)
ggcoefstats(mod_ba_adult)

# models mediators - outcomes
mod_hta   <- stats::lm(formula = hta ~ age*sex + kdm_advance, data = dataBA)
mod_diabetes  <- stats::lm(formula = diabetes ~ age*sex + kdm_advance, data = dataBA)
mod_cvd  <- stats::lm(formula = cvd ~ age*sex + kdm_advance, data = dataBA)
mod_bmi  <- stats::lm(formula = bmi ~ age*sex + kdm_advance, data = dataBA)

ggcoefstats(mod_hta)
ggcoefstats(mod_diabetes)
ggcoefstats(mod_cvd)
ggcoefstats(mod_bmi)

# models exposures - outcomes
mod_hta_aces   <- stats::lm(formula = hta ~ age*sex + household_violence + migration_yo + Class + childhood_sep, data = dataBA)
mod_dm_aces   <- stats::lm(formula = diabetes ~ age*sex + household_violence + migration_yo + Class + childhood_sep, data = dataBA)
mod_cvd_aces  <- stats::lm(formula = cvd ~ age*sex + household_violence + migration_yo + Class + childhood_sep, data = dataBA)
mod_bmi_aces  <- stats::lm(formula = bmi ~ age*sex + household_violence + migration_yo + Class + childhood_sep, data = dataBA)

ggcoefstats(mod_hta_aces)
ggcoefstats(mod_dm_aces)
ggcoefstats(mod_cvd_aces)
ggcoefstats(mod_bmi_aces)


# models mediators - outcomes confounder (health risks)
mod_hta_smo   <- stats::lm(formula = hta ~ age*sex + smoking + alcohol, data = dataBA)
mod_dm_smo   <- stats::lm(formula = diabetes ~ age*sex + smoking + alcohol, data = dataBA)
mod_cvd_smo  <- stats::lm(formula = cvd ~ age*sex + smoking + alcohol, data = dataBA)
mod_bmi_smo  <- stats::lm(formula = bmi ~ age*sex + smoking + alcohol, data = dataBA)

ggcoefstats(mod_hta_smo)
ggcoefstats(mod_dm_smo)
ggcoefstats(mod_cvd_smo)
ggcoefstats(mod_bmi_smo)

# models mediators - outcomes confounder (adult SEP)
mod_hta_adu   <- stats::lm(formula = hta ~ age*sex + education + sep_occupation + current_sep, data = dataBA)
mod_dm_adu   <- stats::lm(formula = diabetes ~ age*sex + education + sep_occupation + current_sep, data = dataBA)
mod_cvd_adu  <- stats::lm(formula = cvd ~ age*sex + education + sep_occupation + current_sep, data = dataBA)
mod_bmi_adu  <- stats::lm(formula = bmi ~ age*sex + education + sep_occupation + current_sep, data = dataBA)

ggcoefstats(mod_hta_adu)
ggcoefstats(mod_dm_adu)
ggcoefstats(mod_cvd_adu)
ggcoefstats(mod_bmi_adu)
```

Check coefficients and R squares exposures-mediators
```{r}
#Exposures-mediators
summary(mod_ba_child)
summary(mod_ba_aces)
summary(mod_ba_adult)

#Mediators-outcomes
summary(mod_hta)
summary(mod_diabetes)
summary(mod_cvd)
summary(mod_bmi)

#Exposures-outcomes
summary(mod_hta_aces)
summary(mod_dm_aces)
summary(mod_cvd_aces)
summary(mod_bmi_aces)

#Mediators-outcomes confounder (health risk)
summary(mod_hta_smo)
summary(mod_dm_smo)
summary(mod_cvd_smo)
summary(mod_bmi_smo)

#Mediators-outcomes confounder (adult SEP)
summary(mod_hta_adu)
summary(mod_dm_adu)
summary(mod_cvd_adu)
summary(mod_bmi_adu)
```

View(dataBA)
```{r}
dataBA$poor_health2 <- as.factor(as.character(dataBA$poor_health2))
```



Check linear regressions models with 
```{r}
set.seed(123)

## models diseases applying DAG confounder-mediators-outcomes
mod_dm   <- stats::lm(formula = kdm_advance ~ age*sex + kdm + hd + hd_log, data = dataBA)
mod_hta  <- stats::lm(formula = kdm_advance ~ age*sex + household_violence + neglected_food + early_infection + poor_health2 + emotional_abuse + migration_yo, data = dataBA)
mod_cvd  <- stats::lm(formula = kdm_advance ~ age*sex + health_insurance + social.cash.transfer + nutrition_program, data = dataBA)
mod_bmi  <- stats::lm(formula = kdm_advance ~ age*sex + smoking + alcohol + raised_bmi, data = dataBA)

ggcoefstats(mod_dm)
ggcoefstats(mod_hta)
ggcoefstats(mod_cvd)
ggcoefstats(mod_bmi)

```

Check coefficients and R squares
```{r}
summary(mod_dm)
summary(mod_hta)
summary(mod_cvd)
summary(mod_bmi)
```

```{r}
## models diseases and biomarkers
mod_dm   <- stats::lm(formula = diabetes ~ bmi + waist + ldl + hdl + trig + totchol +hba1c + glucose + dbp + sbp, data = data)
mod_hta  <- stats::lm(formula = hta ~ bmi + waist + ldl + hdl + trig + totchol +hba1c + glucose + dbp + sbp, data = data)
mod_cvd  <- stats::lm(formula = cvd ~ bmi + waist + ldl + hdl + trig + totchol +hba1c + glucose + dbp + sbp, data = data)
mod_bmi  <- stats::lm(formula = bmi ~ bmi + waist + ldl + hdl + trig + totchol +hba1c + glucose + dbp + sbp, data = data)

ggcoefstats(mod_dm)
ggcoefstats(mod_hta)
ggcoefstats(mod_cvd)
ggcoefstats(mod_bmi)
```


```{r}
mod_childsep <- stats::lm(formula = bmi ~ Class + childhood_sep, data = data)
mod_aduldsep <- stats::lm(formula = bmi ~ education + current_sep + sep_occupation, data = data)
mod_policy  <- stats::lm(formula = bmi ~ health_insurance + social.cash.transfer + nutrition_program, data = data)

ggcoefstats(mod_childsep)
ggcoefstats(mod_aduldsep)
ggcoefstats(mod_policy)

summary(mod_childsep)
summary(mod_aduldsep)
summary(mod_policy)
```




#Model BAA and exposures as predictors
```{r}
mod_markers <- stats::lm(formula = kdm_advance ~ bmi + waist + ldl + hdl + trig + totchol +hba1c + glucose + dbp + sbp , data = data)
mod_baa   <- stats::lm(formula = kdm_advance ~ age + hd + hd_log + kdm, data = data)
mod_aces  <- stats::lm(formula = kdm_advance ~  household_violence + neglected_food + early_infection + poor_health2 + emotional_abuse + migration_yo, data = data)
mod_childsep <- stats::lm(formula = kdm_advance ~ Class + childhood_sep, data = data)
mod_aduldsep <- stats::lm(formula = kdm_advance ~ education + current_sep + sep_occupation, data = data)
mod_policy  <- stats::lm(formula = kdm_advance ~ health_insurance + social.cash.transfer + nutrition_program, data = data)
mod_outcome <- stats::lm(formula = kdm_advance ~ diabetes + hta + cvd + raised_bmi, data = data)
mod_risk    <- stats::lm(formula = kdm_advance ~ alcohol + smoking + medication, data = data)


ggcoefstats(mod_markers)
ggcoefstats(mod_baa)
ggcoefstats(mod_aces)
ggcoefstats(mod_childsep)
ggcoefstats(mod_aduldsep)
ggcoefstats(mod_policy)
ggcoefstats(mod_outcome)
ggcoefstats(mod_risk)
```


```




Regressions BA and exposures

## models diseases
mod_markers <- stats::lm(formula = kdm_advance ~ bmi + waist + ldl + hdl + trig + totchol +hba1c + glucose + dbp + sbp , data = data)
mod_baa   <- stats::lm(formula = kdm_advance ~ age + hd + hd_log + kdm, data = data)
mod_aces  <- stats::lm(formula = kdm_advance ~  household_violence + neglected_food + early_infection + poor_health2 + emotional_abuse + migration_yo, data = data)
mod_childsep <- stats::lm(formula = kdm_advance ~ Class + childhood_sep, data = data)
mod_aduldsep <- stats::lm(formula = kdm_advance ~ education + current_sep + sep_occupation, data = data)
mod_policy  <- stats::lm(formula = kdm_advance ~ health_insurance + social.cash.transfer + nutrition_program, data = data)
mod_outcome <- stats::lm(formula = kdm_advance ~ diabetes + hta + cvd + raised_bmi, data = data)
mod_risk    <- stats::lm(formula = kdm_advance ~ alcohol + smoking + medication, data = data)


ggcoefstats(mod_markers)
ggcoefstats(mod_baa)
ggcoefstats(mod_aces)
ggcoefstats(mod_childsep)
ggcoefstats(mod_aduldsep)
ggcoefstats(mod_policy)
ggcoefstats(mod_outcome)
ggcoefstats(mod_risk)


Check coefficients and R squares
```{r}
summary(mod_markers)
summary(mod_baa)
summary(mod_aces)
summary(mod_childsep)
summary(mod_aduldsep)
summary(mod_policy)
summary(mod_outcome)
summary(mod_risk)

plot(mod_markers)
plot(mod_baa)
plot(mod_aces)
plot(mod_childsep)
plot(mod_aduldsep)
plot(mod_policy)
plot(mod_outcome)
plot(mod_risk)

```
Annotations:
- leave dbp and wc, otherwise model lost Adjusted R-squared and increased Residual standard error. Model markers OK.
- Model baa: fit.
- Model mod_aces: only significant domestic violence and migration. Both significant but explain less Adjusted R-squared.
- mod_childsep: keep CLass child SEP not significant


#significant
```{r}
## models
mod_exposure  <- stats::lm(formula = kdm_advance ~  household_violence + Class + childhood_sep + migration_yo, data = data)
ggcoefstats(mod_exposure)

mod_kdm<- stats::lm(formula = kdm_advance ~ age + race + sex , data = data)
ggcoefstats(mod_kdm)

mod_outcome <- stats::lm(formula = kdm_advance ~ diabetes + hta + cvd + raised_bmi, data = data)
ggcoefstats(mod_outcome)

```


Check coefficients and R squares
```{r}
summary(mod_exposure)
summary(mod_kdm)
summary(mod_outcome)

plot(mod_exposure)
plot(mod_kdm)
plot(mod_outcome)
```

Assessing the Goodness of Fit of the Model

To assess how “good” the regression model fits the data, we can look at a couple different metrics:

1. Multiple R-Squared

This  measures the strength of the linear relationship between the predictor variables and the response variable. A multiple R-squared of 1 indicates a perfect linear relationship while a multiple R-squared of 0 indicates no linear relationship whatsoever.

Multiple R is also the square root of R-squared, which is the proportion of the variance in the response variable that can be explained by the predictor variables. In this example, the multiple R-squared is 0.775. Thus, the R-squared is 0.7752 = 0.601. This indicates that 60.1% of the variance in mpg can be explained by the predictors in the model.



2. Residual Standard Error

This measures the average distance that the observed values fall from the regression line. In this example, the observed values fall an average of 3.008 units from the regression line.

****






```{r}
plot(mod_dm)
plot(mod_hta)
plot(mod_cvd)
summary(mod_bmi)
```





Using the Model to Make Predictions

From the output of the model we know that the fitted multiple linear regression equation is as follows:



Assessing Model Fit:

In typical linear regression, we use R2 as a way to assess how well a model fits the data. This number ranges from 0 to 1, with higher values indicating better model fit.

However, there is no such R2 value for logistic regression. Instead, we can compute a metric known as McFadden’s R2, which ranges from 0 to just under 1. Values close to 0 indicate that the model has no predictive power. In practice, values over 0.40 indicate that a model fits the data very well.

A value of 0.4728807 is quite high for McFadden’s R2, which indicates that our model fits the data very well and has high predictive power.

We can compute McFadden’s R2 for our model using the pR2 function from the pscl package:
```{r}
pscl::pR2(mod_markers)["McFadden"]
pscl::pR2(mod_baa)["McFadden"]
pscl::pR2(mod_aces)["McFadden"]
pscl::pR2(mod_childsep)["McFadden"]
pscl::pR2(mod_aduldsep)["McFadden"]
pscl::pR2(mod_policy)["McFadden"]
pscl::pR2(mod_outcome)["McFadden"]
pscl::pR2(mod_risk)["McFadden"]

pscl::pR2(mod_baa)["McFadden"]
pscl::pR2(mod_baa2)["McFadden"]
pscl::pR2(mod_exposure)["McFadden"]
pscl::pR2(mod_outcome)["McFadden"]

```

Variable Importance:

We can also compute the importance of each predictor variable in the model by using the varImp function from the caret package:

install.packages('caret')
install.packages("caret", dependencies = c("Depends", "Suggests"))
pak::pak('topepo/caret/pkg/caret')
library(package = "lattice")
The caret package (short for Classification And REgression Training) is a set of functions that attempt to streamline the process for creating predictive models.

```{r}
caret::varImp(mod_baa)
caret::varImp(mod_baa2)
caret::varImp(mod_exposure)
caret::varImp(mod_outcome)
```

